---
title: "Visualizing Iowa county populations by year"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, results = 'hide', message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(sf)
library(janitor)
library(tidyverse)
library(tmap)
library(lubridate)
library(gganimate)
```



Data sources:

* [City populations by county and year in Ames from data.iowa.gov](https://data.iowa.gov/Community-Demographics/City-Population-in-Iowa-by-County-and-Year/y8va-rhk)
* [County boundaries of Iowa: geodata.iowa.gov](https://geodata.iowa.gov/dataset/county-boundaries-iowa)


```{r}
iowa.sf <- st_read('data/county') %>% 
  clean_names() %>% 
  st_simplify(dTolerance = 500)

month_to_number <- function(x) {
  x <- tolower(substr(x, 1, 3))
  match(tolower(x), tolower(month.abb))
}

county_pops <- read_csv('data/City_Population_in_Iowa_by_County_and_Year.csv') %>% 
  clean_names() %>% 
  separate('year', c('month', 'day', 'year'), sep = ' ') %>% 
  mutate(year = as.integer(year),
         month = month_to_number(month),
         day = as.integer(day),
         estimate = as.integer(estimate),
         county = replace(county, county == "O'Brien", "Obrien"))

```

Summarizing the county population data by year and county so that we can join it to `iowa.sf`. We also add a column for percentage change over last year because population stays roughly the same within counties, which doesn't make for very interesting graphs.

```{r}
county_by_year <- county_pops %>% 
  group_by(county, year) %>%  
  summarise(total_pop = sum(estimate, na.rm = TRUE)) %>% 
  mutate(lag = lag(total_pop),
         pct.change = (total_pop - lag) / lag) 
```

Adding county populations to the iowa.sf object.

```{r}
iowa.sf <- inner_join(iowa.sf, county_by_year, by = 'county') 
```

Visualizing counties and population using tmap.

```{r}
iowa.sf %>% 
  tm_shape() + 
  tm_fill(
    col = 'total_pop',
    title = "Total Population (2018)"
  ) +
  tm_borders(lwd = 0.5) +
  tm_text('county', size = 0.6) + 
  tm_layout(
    "Ames Counties",
    inner.margins = c(0.08, 0.08, 0.08, 0.08),
    legend.position = c('left', 'bottom'),
    legend.title.size = 1,
    title.position = c("center", "top"),
  ) + 
  tm_credits("Source: https://data.iowa.gov/Community-Demographics/City-Population-in-Iowa-by-County-and-Year/y8va-rhk9",
             position = c(0.37, 0.0))
```

Animation of total population over 2011 - 2018. We drop 2010 because there is no percentage change data for that year.

```{r}
iowa.sf %>% 
  filter(year > 2010) %>% 
  ggplot(aes(fill = total_pop)) +
  geom_sf() + 
  ggthemes::theme_map() + 
  labs(
    title = "Ames population by county", 
    subtitle = "Year: { current_frame }",
    caption = "Source: https://data.iowa.gov/Community-Demographics/City-Population-in-Iowa-by-County-and-Year/y8va-rhk9"
  ) + 
  scale_fill_gradient2(low = 'blue', mid = 'white', high = 'red') +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    legend.title = element_blank(),
    legend.background = element_rect(colour = 'black', fill = 'white'),
    panel.background = element_rect(fill = 'gray95')
  ) + 
  transition_manual(year) +
  geom_sf_text(aes(label = county), size = 2.25)

```

Not very exciting because population stays pretty constant over years:

```{r}
iowa.sf %>% 
  filter(county %in% sample(unique(iowa.sf$county), 20)) %>% 
  ggplot(aes(year, total_pop)) +
  geom_col() + 
  facet_wrap(~ county, scales = "free_y") + 
  theme(axis.text.y = element_blank(),
        axis.text.x = element_text(angle = 90)) +
  labs(x = element_blank(),
       title = "Population in Ames counties from 2011 to 2018"
       ) + 
  scale_x_continuous(breaks = 2010:2018)
```

Mapping percentage change instead:

```{r}

iowa.sf %>% 
  filter(year > 2010) %>% 
  ggplot(aes(fill = pct.change)) +
  geom_sf() + 
  ggthemes::theme_map() + 
  labs(title = "Percent change in Ames population by county", 
       subtitle = "Year: { current_frame }",
       caption = "Source: https://data.iowa.gov/Community-Demographics/City-Population-in-Iowa-by-County-and-Year/y8va-rhk9"
       ) + 
  scale_fill_gradient2(low = 'blue', mid = 'white', high = 'red',
                       na.value = 'white'
                       ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    legend.title = element_blank(),
    legend.background = element_rect(colour = 'black', fill = 'white'),
    panel.background = element_rect(fill = 'gray95')
  ) + 
  transition_manual(year) +
  geom_sf_text(aes(label = county), size = 2.25) 

```

We can see by the scale that the percentage change isn't very large, but at least the animation is more interesting :)
